---
title: "Lab 03"
author: "Dr. Xiang Ji @ Tulane University"
date: "Feb 25, 2022"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: Math 6040/7260 Linear Models
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
rm(list = ls()) # clean-up workspace
```

## Simple Linear Regression

The `Davis` data set in the **carData** package contains the measured and self-reported heights and weights of 200 men and women engaged in regular exercise.  A few of the data values are missing, and consequently there are only 183 complete cases for the variables that are used in the analysis reported below.

First, take a quick look at the data:

```{r}
library(carData)
summary(Davis)
```

### Q1. How many variables are in the `Davis` data and explain what they represent


We focus here on the regression of `weight` on `repwt`.  This problem has response $y=weight$ and one predictor, `repwt`, from which we obtain the regressor $x_1=repwt$.  We again construct the design matrix and response vector first.

```{r}
X <- as.matrix(cbind(1, Davis[, "repwt"]))
Y <- as.matrix(Davis[, "weight"])

# Now get X rows with complete observations
X.complete.rows <- complete.cases(X)  # read the documentation of funciton complete.cases()
# Similarly get Y's
Y.complete.rows <- complete.cases(Y)

# subset X and Y 
X <- X[X.complete.rows & Y.complete.rows, ]
Y <- Y[X.complete.rows & Y.complete.rows, ]
```

### Q2.  Calculate the least square estimate using $\hat{\beta_{ls}} = (X^TX)^{-1}X^TY$.

```{r}

```

### Q3. Calculate the total sum of squares, residual sum of squares and regression sum of squares.  Test $SST - SSR - SSReg$, explain what you get.

```{r}

```

### Q4. Calculate the mean squared error (MSE).

```{r}

```

### Q5. Calculate the test statistic for testing the null hypothesis $H_0: \beta_0 = 0$ vs $H_a: \beta_0 \ne 0$, report the p-value.

```{r}

```

### Q6. Fit the same model using `lm()` function and check your calculations above

```{r}

```

### Q7. Repeat Q2 to Q6 for the dataset `Mandel` to regress `y` on `x1`.

```{r}

```
