---
title: "Lecture 32: Lab 09"
author: "Dr. Xiang Ji @ Tulane University"
date: "April 22, 2022"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-6040/7260 Linear Models
csl: ../apa.csl
---

## Q.1 Sample size calculation for one-way ANOVA model

Let's now revisit the binding fraction example for sample size calculation.
Suppose that we want to test equal mean binding fractions among antibiotics against the alternative
$$
H_1: \mu_p = \mu + 3, \mu_t = \mu+3, \mu_s = \mu - 6, \mu_E = \mu, \mu_C = \mu
$$
so that
$$
\mu_ 1 = \mu_2 = 3, \mu_3 = -6, \mu_4 = \mu_5 = 0.
$$

Assume $\sigma = 3$ and we need to use $\alpha = 0.05$.
The noncentrality parameter is given by
$$
\gamma = n[(\frac{3}{3})^2 +(\frac{3}{3})^2 + (\frac{-6}{3})^2]
$$
and the $\alpha = 0.05$ critical value for $H_0$ is given by
$$
F^* = F(5 - 1, 5n - 5, 0.05)
$$
We need the area to the right of $F^*$ for the noncentral $F$ distribution with degrees of freedom $4$ and $5(n - 1)$ and noncentrality parameter $\gamma = 6n$ to be greater or equal to the desired power level of $1 - \beta = 0.8$.

Now we will do the "real" calculations to finish this example.

We proceed by filling out the columns of the following table step by step.

| $n$ | ndf | ddf | NCP | QF | PF | POWER |
|:---:|:-----:|:-----:|:---:|:--:|:--:|:-----:|
|2 | | | | | | |
|3 | | | | | | |
|4 | | | | | | |
|5 | | | | | | |
|6 | | | | | | |
|7 | | | | | | |
|8 | | | | | | |
|9 | | | | | | |
|10 | | | | | | |

### a. prepare data

```{r}
rm(list=ls())  # clean up workspace
library(tidyverse)
binding.fraction.data <- matrix(
  c(29.6, 24.3, 28.5, 32.0,
		 27.3, 32.6, 30.8, 34.8,
		 5.8, 6.2, 11.0, 8.3,
		 21.6, 17.4, 18.3, 19,
		 29.2, 32.8, 25.0, 24.2), nrow = 5, ncol = 4, byrow = TRUE
)
rownames(binding.fraction.data) <- c("Penicillin G", "Tetracyclin", "Streptomycin", "Erythromycin", "Chloramphenicol")

binding.df <- as_tibble(binding.fraction.data, rownames = NA) %>%
  rownames_to_column(var = "antibiotic") %>%
  pivot_longer(cols = 2:5, values_to = "bf", names_to = NULL)
```

Let's consider a sequence of $n$ values from $2$ to $10$.

```{r}
n <- 2:10
```

### b. fill out the column of denominator df
```{r}
```

### c. fill out the column of noncentrality parameter (NCP)

```{r}
```

### d. fill out the column of critical value $F^*$ (QF)

```{r}
```

### e. fill out the column of area to the **left** of $F^*$ (PF)

$$
 \Pr(MS[Trt]/MS[E] < F^*; H_1 \mbox{ is true})
$$

```{r}
```

### f. fill out the column of power (which is the **right** of $F^*$ under $H_1$)
$$
 \Pr(MS[Trt]/MS[E] > F^*; H_1 \mbox{ is true})
$$

```{r}
```

### g. what does the table look like now?



### h. what is the lowest sample size that obtains at least power of $0.8$?


## Q.2 Use the function `power.anova.test`

Read the documentation of `power.anova.test` function in R.
You may use this webpage for reference [webpage link](https://stats.idre.ucla.edu/r/dae/one-way-anova-power-analysis/).

Repeat the power calculations above using `power.anova.test` function.  Do you obtain the same values?

```{r}
```








